---
title: "5 Minute Tutorials: Guarding Against Spatial *and* Temporal Leakage"
author: 'Mike Gaunt'
date: '2025-04-11'
slug: spatiotemporal-cv
categories: ["R", "Spatiotemporal Modeling"]
tags: ["spatial CV", "temporal CV", "rsample", "spatialsample", "custom resampling"]
description: "A custom resampling strategy that prevents both spatial and temporal leakage — because modeling isn’t magic, it’s just well-split data."
comments: yes
image: ~
math: ~
license: ~
hidden: no
---

# Quick Overview

Hey hey —  

Today’s 5-minute trick tackles a big gap I kept running into: 

**I needed to model something with both spatial structure _and_ time dependencies**, but nothing I found in R’s modeling universe would let me **resample across both space and time** _without leakage_.

## What's the modeling task?

Let’s say you want to **spatially interpolate predictions** (i.e., predict at new locations) **_and_ forecast in time** (i.e., predict future values). Your model has to generalize both *in space* and *in time*.

If you don’t split your data smartly — say you train your model on one time slice and evaluate on the same locations a few days later — you might **leak information** across your train/test sets and falsely inflate your model's performance.

Even worse, if you’re tuning a model and evaluating performance with leakage, you're optimizing toward something that will crumble in production.

In order to do this, I wrote my own function that integrates with and extends the tidymodels' existing sampling framework  
And yes, I’m going to share it 😎

---

## Required Libraries

```{r results='hide', message=FALSE, warning=FALSE}
library(CAST)
library(tidyverse)
library(sp)
library(sf)
library(rsample)
library(plotly)
library(crosstalk)
```

---

# What We're Modeling

We're working with the `cookfarm` dataset from the `CAST` package.  
 
Each record is a repeated measurement (e.g., soil moisture) from one of many **spatial sensor locations** over time. So the dataset is **spatiotemporal**, with groupings by sensor.

Here's a peek:

```{r}
data("cookfarm")
glimpse(cookfarm)
```

Goals:

- Split the data **spatially** so no sensors (SOURCEIDs) appear in both train and test
- Then **temporally** within each spatial fold using a sliding window

Leakage would allow the model to "peek" at test data by sharing sensors or future observations between train/test - which is bad!!!.

---

# Step 1: Spatial Folds

We'll take a slice of the data (two weeks in 2012 at one depth), and convert it to an `sf` object:

```{r}
trainDat <- cookfarm %>%
  filter(altitude == -0.3,
         lubridate::year(Date) == 2012,
         lubridate::week(Date) %in% 13:14) %>%
  mutate(Date = as.Date(Date)) %>%
  st_as_sf(coords = c("Easting", "Northing"), crs = 26911)
```

Now we split spatially so each fold has unique sensors (`SOURCEID`s). 

You could also use `spatial_clustering_cv()` from the `spatialsample` package — that’s a great option too.

```{r}
spatial_folds <- trainDat %>%
  group_vfold_cv(group = SOURCEID, v = 5)

spatial_folds
```

Now we have training and validation folds that are **mutually exclusive in space**.

Next up: **time**.

---

# Step 2: Rolling Origin… with Groups?!

Rolling origin resampling (`sliding_window()`, etc.) is great for time —  
but it fails with grouped data, like multiple sensors. 

It also expects a singular time signal so it just slices the row indices, ignoring your groupings.

So I wrote a helper that loops through each sensor group, applies rolling origin resampling, and merges the results:

### `create_grouped_rolling_origin()`:

```{r}
create_grouped_rolling_origin <- function(data, initial, assess, cumulative = TRUE, skip = 0) {
  sensors <- unique(data$SOURCEID)
  splits_list <- list()
  i = 0
  
  for (sensor in sensors) {
    i = i + 1
    sensor_data <- data %>% 
      filter(SOURCEID == sensor) %>%
      arrange(Date)

    if (nrow(sensor_data) >= (initial + assess)) {
      ro <- rolling_origin(sensor_data, 
                           initial = initial, 
                           assess = assess, 
                           cumulative = cumulative, 
                           skip = skip) %>% 
        mutate(sensor = sensor)

      splits_list[[i]] = ro
    }
  }
    
  splits_list = splits_list %>% 
    reduce(bind_rows) %>%  
    group_by(id_1 = id) %>% 
    group_map(~{
      temp_analysis = .x$splits %>% 
        map_df(~ analysis(.x))
      
      temp_assessment = .x$splits %>% 
        map_df(~ assessment(.x))
      
      combined_data <- bind_rows(temp_analysis, temp_assessment)
      n_analysis <- nrow(temp_analysis)
      
      structure(
        list(
          data = combined_data,
          in_id = 1:n_analysis,
          out_id = (n_analysis + 1):nrow(combined_data)
        ),
        class = c("sliding_period_split", "rsplit")
      )
    })
  
  inner_rset <- rsample::new_rset(
    splits = splits_list,
    ids = paste0("Resample", seq_along(splits_list))
  )
  
  class(inner_rset) <- c("sliding_period", "rset", "tbl_df", "tbl", "data.frame")
  
  return(inner_rset)
}
```

Apply it to one spatial fold:

```{r}
example_fold = spatial_folds$splits[[1]] %>% analysis()

inner_resamples = create_grouped_rolling_origin(
  data = example_fold,
  initial = 8,
  assess = 2,
  cumulative = TRUE)
```

---

# Step 3: Peek at the Splits Over Time

Let’s **visualize the rolling windows** per sensor.
    
Each plot shows which days were in train/test sets per sensor.

Helper function:

```{r}
view_split_groups_date <- function(all, train, val){
  bind_rows(
    all %>%
      count(Date) %>%
      mutate(SOURCEID = "All Data", type = "all"),
    train %>%
      count(SOURCEID, Date) %>%
      mutate(type = "Train"),
    val %>%
      count(SOURCEID, Date) %>%
      mutate(type = "Validation")
  )
}
```

Now we make a `plotly` interactive plot:

```{r}
data = list(
  inner_resamples$splits,
  1:length(inner_resamples$splits)
) %>% 
  pmap(~{
    view_split_groups_date(
      example_fold %>% st_drop_geometry(),
      .x %>% analysis(),
      .x %>% assessment()
    ) %>% 
      mutate(sample = str_glue("Sample_{.y}"))
  }) %>%  
  reduce(bind_rows) %>% 
  mutate(type_num = as.factor(type) %>% as.numeric()) %>%  
  mutate(text = str_glue("Sensor: {SOURCEID}\nDate: {Date}\nType: {type}"))

data_sd = SharedData$new(data)

bscols(
  widths = c(4, 12),
  crosstalk::filter_checkbox(
    "id", "Inner Sliding Window Sample", data_sd,
    ~sample, inline = T, allLevels = F),
  data_sd %>% 
    plot_ly(
      x = ~Date, y = ~SOURCEID, z = ~type_num,
      type = "heatmap", text = ~text, hoverinfo = "text"
    ) %>% 
    layout(
      yaxis = list(title = "Sensor Locations"), xaxis = list(title = "")) %>% 
    hide_colorbar()
)
```

Check it! Each inner sample (time samples) depicts sliding windows of train and validation sets respective for each sensor.

---

# Why Bother?

This might feel like a lot, but it’s one of those cases where **doing it right matters**:

- Most tools only resample in time _or_ space — **not both** (and def not when there's repeated measures)
- This custom approach avoids data leakage in **both dimensions**
- Great for **spatiotemporal interpolation and forecasting**

And: if you're tuning a model with `tune_grid()` or `workflow_map()`, **nesting this inside your outer spatial folds** helps ensure leakage doesn't sneak in during hyperparameter selection — especially with temporal dependencies in the data.

So: split smart.  
Model well.  
Avoid leakage.  
