---
title: "5 Minute Tutorials: Guarding Against Spatial *and* Temporal Leakage"
author: 'Mike Gaunt'
date: '2025-04-11'
slug: spatiotemporal-cv
categories: ["R", "Spatiotemporal Modeling"]
tags: ["spatial CV", "temporal CV", "rsample", "spatialsample", "custom resampling"]
description: "A custom resampling strategy that prevents both spatial and temporal leakage — because modeling isn’t magic, it’s just well-split data."
comments: yes
image: ~
math: ~
license: ~
hidden: no
---


# Quick Overview

Hey hey —  
Today’s 5-minute trick tackles a big gap I kept running into: **I needed to model something with both spatial structure _and_ time dependencies**, but nothing I found in R’s modeling universe would let me **resample across both space and time** _without leakage_.

So I wrote my own function for it.  
And yes, I’m going to share it 😎

---

Required libraries.

```{r}
library(CAST)
library(tidyverse)
library(sp)
library(sf)
library(rsample)
library(plotly)
library(crosstalk)
```

# What We're Modeling

We're working with the `cookfarm` dataset from the `CAST` package. Think of it as repeat measurements (e.g., soil moisture) across a bunch of **spatial sensor locations** over time. Here’s a tiny peek:

```{r}
data("cookfarm")
cookfarm %>%
  arrange(SOURCEID, Date) %>%
  head(10)
```

Each `SOURCEID` is a different sensor. We want to:

- Split the data **spatially** so no sensors appear in both training and test sets,
- Then split each of those folds **temporally** using a sliding window.

Why? Because **leakage ruins modeling** — and this guards against both spatial and temporal flavors of it.

---

# Step 1: Spatial Folds

We start by selecting a small slice of data (two weeks in 2012 at one depth), then convert it to an `sf` object:

```{r}
trainDat <- cookfarm %>%
  filter(altitude == -0.3,
         lubridate::year(Date) == 2012,
         lubridate::week(Date) %in% 13:14) %>%
  mutate(Date = as.Date(Date)) %>%
  st_as_sf(coords = c("Easting", "Northing"), crs = 26911)

trainDat %>% glimpse()
```

Now we split spatially so each fold includes unique sensors (`SOURCEID`s):

```{r}
spatial_folds <- trainDat %>%
  group_vfold_cv(group = SOURCEID, v = 5)

spatial_folds
```

Now we’ve got training and validation folds that are **mutually exclusive spatially**. Next up: **time**.

---

# Step 2: Rolling Origin… with Groups?!

Rolling origin resampling (`sliding_window()`, etc.) is awesome — but it **fails** when your data has groups (like different sensors). It just slices based on the data frame index. So I wrote a helper function that loops through each sensor group, creates rolling origin resamples within each group, and then merges the slices.

### Here's the function:
(You can paste in your `create_grouped_rolling_origin()` function here)
```{r}
create_grouped_rolling_origin <- function(data, initial, assess, cumulative = TRUE, skip = 0) {
  # browser()
  sensors <- unique(data$SOURCEID)
  splits_list <- list()
  i = 0
  
  for (sensor in sensors) {
    i = i + 1
    sensor_data <- data %>% 
      filter(SOURCEID == sensor) %>%
      arrange(Date)
    # Only create splits if there are enough rows (at least initial + assess)
    if (nrow(sensor_data) >= (initial + assess)) {
      ro <- rolling_origin(sensor_data, 
                           initial = initial, 
                           assess = assess, 
                           cumulative = cumulative, 
                           skip = skip) %>% 
        mutate(sensor = sensor)
      # Append all splits from this sensor to the list
      # splits_list = list(splits_list, ro)
      splits_list[[i]] = ro
    }
  }
    
    splits_list = splits_list %>% 
      reduce(bind_rows) %>%  
      group_by(id_1 = id) %>% 
      group_map(~{
        temp_analysis = .x$splits %>% 
          map_df(~{
            analysis(.x)
          })
        
        temp_assessment = .x$splits %>% 
          map_df(~{
            assessment(.x)
          })
        
        combined_data <- bind_rows(temp_analysis, temp_assessment)
        n_analysis <- nrow(temp_analysis)
        
        my_rsplit <- structure(
          list(
            data = combined_data,
            in_id = 1:n_analysis,
            out_id = (n_analysis + 1):nrow(combined_data)
          ),
          class = c("sliding_period_split", "rsplit")
          
        )
        
      })
    
    inner_rset <- rsample::new_rset(
      splits = splits_list,
      ids = paste0("Resample", seq_along(splits_list))
    )
  
    class(inner_rset) <- c("sliding_period", "rset", "tbl_df", "tbl", "data.frame")
  
  return(inner_rset)
}

```


Let’s apply it to just one spatial fold:

```{r}
example_fold <- spatial_folds$splits[[1]] %>% analysis()
inner_resamples <- create_grouped_rolling_origin(
  data = example_fold,
  initial = 10,
  assess = 2,
  cumulative = TRUE
)
```

---

# Step 3: Peek at the Splits Over Time

Let’s **visualize the rolling time windows** per sensor.  
Each plot will show which days are in the training vs assessment sets for each sensor.

```{r}
# helper function
view_split_groups_date <- function(all, train, val){
  bind_rows(
    all %>%
      count(Date) %>%
      mutate(SOURCEID = "All Data", type = "all"),
    train %>%
      count(SOURCEID, Date) %>%
      mutate(type = "train"),
    val %>%
      count(SOURCEID, Date) %>%
      mutate(type = "val")
  )
}
```

Now we generate a **step-through plot** using `plotly`:

```{r}

data = list(
  inner_resamples$splits
  ,1:length(inner_resamples$splits)
) %>% 
  pmap(~{
    
    # browser()
    view_split_groups_date(
      temp_all
      ,.x %>% analysis()
      ,.x %>% assessment()
    )   %>% 
      mutate(sample = str_glue("Sample_{.y}"))
      
  }) %>%  
  reduce(bind_rows) %>% 
  mutate(type_num = as.factor(type) %>% 
           as.numeric()) %>%  
  mutate(text = str_glue("Sensor: {SOURCEID}\nDate: {Date}\nType: {type}"))

library(crosstalk)
data_sd = SharedData$new(data)

bscols(
  widths = c(4, 12)
  ,crosstalk::filter_checkbox("id", "Inner Sliding Window Sample", data_sd, ~sample, inline = T, allLevels = F)
  ,data_sd %>% 
    plot_ly(
      x = ~Date,
      y = ~SOURCEID,
      z = ~type_num,
      type = "heatmap",
      text = ~text, hoverinfo = "text"
      
    ) %>% 
    layout(
      yaxis = list(title = "Sensor Locations")
      ,xaxis = list(title = "") ) %>% 
    hide_colorbar()
)


```

---

# Why Bother?

This might feel like a lot, but honestly:

- Most tools only resample in time _or_ space — **not both**.
- This avoids data leakage while giving you proper spatiotemporal CV.
- It’s a flexible pattern: plug in whatever time window or group you want.

You could even nest this inside `tidymodels::workflow_sets()` if you’re into that life.

Enjoy! 🧠🛰️📆
