---
title: "5 Minute Tutorials: Tricks with Purrr::map() #1"
author: 'Mike Gaunt'
date: '2023-10-13'
slug: quickie-purrr
categories: ["R", "Programming"]
tags: ["purrr", "vectorization", '5Min Tutorials']
description: "Use map to use functions on dataframes that you otherwise couldn't!"
image: ~
math: ~
license: ~
hidden: no
comments: yes
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(digest)
```

```{r echo=FALSE, message=FALSE}
apply_digest_to_column <- function(col) {
  map_chr(col, ~digest(object = .x, algo = "crc32", serialize = FALSE))
}
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#this chunk creates all of the objects for the tutorial 
full_string_to_root = "C:/Users/USMG687637/Documents/080_projects/replica_data_requests"

files = list.files(
  full_string_to_root, recursive = TRUE
           ,include.dirs  = T, all.files = T)

# print(head(files))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
temp_data = data.frame(paths = c(files[str_detect(files, "\\.R$")]
                     ,files[str_detect(files, "\\.r$")])) %>% 
  separate(col = "paths", sep = "/", into = paste0("temp_", 1:6), remove  = F) %>% 
  pivot_longer(cols = starts_with("temp")
               ,values_to = "file_name") %>% 
  filter(!is.na(file_name)) %>% 
  mutate(name = parse_number(name)) %>% 
  group_by(paths) %>% 
  filter(name == max(name)) %>% 
  ungroup() %>% 
  mutate(paths = str_remove(paths, paste0("/", file_name))) %>% 
  select(paths, file_name)

temp_data_public =  temp_data %>%  
    mutate(across(c(paths, file_name), apply_digest_to_column))  

example_single = here::here(
  full_string_to_root
  ,temp_data$paths[1]
  ,temp_data$file_name[1]) %>% 
  file.info() 

rownames(example_single) = "example_file_path_for_file"

example_df = temp_data %>% 
  mutate(file_info = here::here(
    full_string_to_root
    ,temp_data$paths
    ,temp_data$file_name) %>% 
      file.info()) %>% 
  mutate(across(c(paths, file_name), apply_digest_to_column))  

# colnames(example_df)
# colnames(example_df$file_info)

example_df_pro =  example_df %>%  
  select(paths, file_name) %>% 
  bind_cols(
    temp_data %>% 
      mutate(info = map(
        here::here(
          full_string_to_root
          ,paths
          ,file_name), file.info)) %>% 
      select(!c(paths, file_name)))

example_df_pro_unnested = example_df_pro %>%  
  unnest(cols = info)

# colnames(example_df_pro_unnested)
```


# Quick Overview

Hi hi!   
This is intended to  be a very quick tutorial highlighting bite-sizeable and really helpful R tips. 

## What are we looking at today

I wanted to share how to use `purrr::map()` to apply function that produces more that one output (columns) to a dataframe.

## How
We ware going to make basic script that lists all files in a directory and its sub-directories that have a specific extension.   
This was something that I needed to do at work and could not, for the life of me, find a file that I was looking for...

## What you need 

Libraries:
```{r}
library(tidyverse)
```

## Base Data

The code below makes the datafame that we will use for this tutorial

```{r eval=FALSE}
full_string_to_root = "<<put a full path to a directory that has a lot of R scripts in it>>"

#get all files in the root
files = list.files(
  full_string_to_root, recursive = TRUE
           ,include.dirs  = T, all.files = T)

#filter files for the root 
#split file paths to separate the file path and name in different columns 
#NOTE: this is wayyyyy more efficient with REGREX but I couldn't make it work
temp_data = data.frame(paths = c(files[str_detect(files, "\\.R$")]
                     ,files[str_detect(files, "\\.r$")])) %>% 
  separate(col = "paths", sep = "/", into = paste0("temp_", 1:6), remove  = F) %>% 
  pivot_longer(cols = starts_with("temp")
               ,values_to = "file_name") %>% 
  filter(!is.na(file_name)) %>% 
  mutate(name = parse_number(name)) %>% 
  group_by(paths) %>% 
  filter(name == max(name)) %>% 
  ungroup() %>% 
  mutate(paths = str_remove(paths, paste0("/", file_name))) %>% 
  select(paths, file_name)

```

# The actual tutorial

So this is what my dataframe of files looks like...

I hashed my files - randomly coded them - since they are sensitive:

Yours should be something to the effect of `code folder` `script.R`. And it doesn't matter how deep your folder tree goes. 

```{r }
print(temp_data_public)
```

So this is cool, we listed all the files below our root director but the names alone don't tell us much. 

To make it more useful, we will add columns indicating when it was made and last modified. 

I'm going to do this by supplying the `file.info()` with a full path of each script in our DF via the `here::here()` function.

It's easy for just one file at a time:
```{r eval=F}
here::here(
  full_string_to_root
  ,temp_data$paths[1]
  ,temp_data$file_name[1]) %>% 
  file.info() 
```

```{r echo = F}
print(example_single)
```

And now we apply this to each record of our DF. 

```{r eval=FALSE}
example_df = temp_data %>% 
  mutate(file_info = here::here(
    full_string_to_root
    ,temp_data$paths
    ,temp_data$file_name) %>% 
      file.info())
```

Take a look at what we just made:
```{r echo=FALSE}
example_df
```

Oops! It looks a little strange, the columns created by the `file.info()` function all have `$s` in front of them. 
```{r}
colnames(example_df)
colnames(example_df$file_info)
```

Seems that we made a `file_info` column that has columns of its own... hmmmmmm.

This is a common problem you can run into this when the function you apply to your DF produces multiple output columns.

# The Fix

The fix is rather simple, we will use `purrr::map()` to effectively make a nested column (much like the DF above) and unnest it to create a _normal_ looking DF.

In essence, all you do is:

+ add a `map()` function in the `mutate()` function
+ supply the function with a column to run throught he function 
  - Note in our case we are sending TWO columns and a string variable but they are collapsed together via the `here::here()` - which is created as one by the `map()`
+ define the function to augment the supplied column with

```{r eval=FALSE}
example_df_pro = temp_data %>% 
  #start of the mutuate()
  mutate(info = map(   #applying the map()
    here::here(
      full_string_to_root
      ,paths
      ,file_name)    #end of the first input, defining what we will map thru
    ,file.info     #defining the function to apply - note: we do not include `()`
  )
  )
```

And it looks like this:
```{r}
example_df_pro
```

Similar to the first DF that we made but we no can see the size of the DF that is in each info record. 

We can expand this by running it through `unnest()`

```{r eval=F}
example_df_pro_unnested = example_df_pro %>%  
  unnest(cols = info)
```

```{r echo=FALSE}
example_df_pro_unnested
```

And you can see that we fixed the column issue....
```{r}
colnames(example_df_pro_unnested)
```


# Parting words

So why do this?!??!? 

For a few reasons: 

+ to be able to deploy functions that have multiple outputs to a dataframe without creating a weird DF structure
+ nesting DFs are a really great way to keep many augmentations to a DF in the same location
  - a lot of novices will do something to a DF and save it as a separate object 
  - this can confuse code, make it difficult to track objects for analysis
  - this one object can be then saved as an RDS object! cool!
  
Good luck and thanks for checking this out!

