---
title: measuring-power
author: ''
date: '2024-11-18'
slug: measuring-power
categories: []
tags: []
description: null
image: null
math: null
license: null
hidden: no
comments: yes
editor_options: 
  markdown: 
    wrap: 72
---

---
title: "Measuring Power in Training: Establishing a Baseline"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(tidyverse)

full_path = here::here("content/post/2024-11-18-measuring-power")
source(
  here::here(full_path, "process_peak_force.R")
)


data = here::here(full_path, "peakload_data_left_right_18_11_2024.csv") %>%
  read.csv() %>% 
  filter(tag == "test_halfcrimp_power")

```


```{r, echo=F}
# Example data filtering and preparation
process_peak_force()
df = data %>%
    filter(str_detect(tag, "power")) %>%
    mutate(
      base_weight = gsub(";.*", "\\1", comment) %>% parse_number(),
      session = gsub(".*(session_)", "\\1", comment)) %>% 
  mutate(across(contains("max.weight"), ~round(.x/base_weight, 2))) %>% 
  pivot_longer(cols = contains("max.weight")
               ,values_to = "accel_g")

# Fit a logarithmic model
model = lm(log(accel_g) ~ log10(base_weight), data = df)

# Create new data for predictions
new_data = data.frame(
  base_weight = seq(15, 150, 5))

# Generate predictions and prediction intervals
au =broom:: augment(model, newdata = new_data, interval = "prediction") %>% 
  mutate(across(c(.fitted:.upper), exp))

# Plot the data and model
df %>% 
  ggplot() + 
  geom_point(aes(base_weight, accel_g)) +
  geom_line(data = au, aes(base_weight, y = .fitted)) + 
  geom_ribbon(data = au, aes(base_weight, ymin = .lower, ymax = .upper), alpha = 0.3) +
  geom_hline(yintercept = 1) + 
  labs(x = "Acgtual Weight Lifted (lbF)", y = "Force Multiplier") + 
  coord_cartesian(x = c(20, 130))
```

# Introduction

Being a boulder bro, I've always focused on being as strong as I can be and have seen it as the one true path to sending nirvana. I'm the guy Sharma screaming on plastic and dyno-ing everything.

This mindset has only gotten worst, as I've been able to gameify my hangboarding session by the use of a TIndeq and introducing the 'deadlifting' method into my handboard session. 

Despite being able to rip 170% body weight in a halftime, I've not been able to some of my big outdoor projects. Seems my issue is on power/contact strength (espially on small 'lil outdoor holds) and not strength. 

Great.... so I now know what the problem is, but how do I go about fixing it and just as importantly, _how do i go about measruing ability and progress_?

This blog post explains the latter, my process in measuring power. I devloped this method in  from a first principabs approach and with the intent of making it as user friednly and easily adoptable. 



# Principles of Power

Power so what is it and how is it different from stength?

Effectively, power is the rate at which strength can be applied, or for those more math literate:

$$
P = F \cdot v
$$

Where: - $P$: Power - $F$: Force applied - $v$: Velocity of movement

A lot of "power" training is facilitated by use of instrunments that measure velcotiy, this type of triaing is called Velocity based training or VBT for short. 

And long story short, i have no wau to directly measure the velocity in a tradtiional context or in a climbing context... but i do have a work around ].

Tindeqs (which i have) measure force, if you slowly lift a 40 LBF wieght it will measure 40 Lbf. 

Now for some physics, fundemenatally: 

Force equals mass times acceleration 

if we acceelerate a mass slowly, just overcoming the acceleration due to gravtiy we will have a overcome (lift) the force of a weight that is more o rless the weight printed on the weight.

HOWEVER, if we lift the a weight (or mass) really, really quickly, we will actually lify physicallhy heavier weight....

Example here::

pls out example of a 
45 lbf weight with accerleation of 9.81 m/s2
and then the mass of that same object being moved at 13.5m/s2 fled compare teh differences in actaul force feld

and then say something about 13.5m/s2 being some multiple of gravity G

This acceleration force


Since velocity is the change in position over time, and acceleration
($a$) is the rate of change of velocity, we can indirectly measure
velocity by analyzing the force required to accelerate a mass over a
short duration.

This principle forms the foundation for training power: the ability to
apply force quickly across a variety of weights.

Unfortunately, there is no free lunch - heavier weights are harder to move than lighter ones (duh.) and in turn, it takes more force to move a given weight faster. 

This relationship is captured in the **force-velocity curve**, which highlights the trade-off between force and velocity during movement.

Why is this important? Contact strength...

Imagine you're protecting a boulder problem. 

The crux is a throw to tenuous crimp where you are forced to cut feet. 

You can dangle on the crimp all day but when you actually are foced to hop and the wall and pefrom the movement, you just can't latch it. 

In either case your fingers must apply a certain amount of wegith to the hold to keep you off the ground.

When you grab the hold and lift your body off the ground, you pefrom this action slowly over a long time window, thus the 

In the instance of the throw - there is a base amount of force you need to apply to the hold but over a very short amount of time (compensating for )

# Methodology

To estimate power without direct velocity measurements: 1. **Force
Measurement**: I measured the maximum force I could generate while
lifting a known weight as fast as possible using consistent grip
positions. 2. **Acceleration Calculation**: I reported acceleration in
terms of **total G's**, calculated as $a = F_{peak} / F_{weight}$, where
$F_{peak}$ is the peak measured force and $F_{weight}$ is the base
weight. 3. **Force-Velocity Relationship**: I plotted force and
estimated acceleration to create a force multiplier curve, showing how
lifting effort changes with mass.

### Example Calculation

To better understand this process, let’s break down an example:

-   **Known Mass (Weight)**: 80 lbf
-   **Measured Peak Force**: 130 lbf
-   **Acceleration Calculation**:

We calculate the total acceleration in terms of G's (multiples of gravities) as:

$$
a_{g} = \frac{F_{peak}}{F_{weight}}
$$

Where: - $F_{peak} = 130 \, \text{lbf}$ -
$F_{weight} = 80 \, \text{lbf}$

Substitute the values:

$$
a_{g} = \frac{130}{80} = 1.63 \, G
$$
This means the total acceleration applied to the base weight is **1.63
G's**.

# Data and Analysis

Here’s how I implemented this analysis in R. The code filters relevant
data, calculates the base weight, and fits a logarithmic model to
explore the relationship between weight and acceleration.

```{r, echo=TRUE}
# Example data filtering and preparation
df = temp_data %>%
    filter(str_detect(tag, "power")) %>%
    mutate(
      base_weight = gsub(";.*", "\\1", comment) %>% parse_number(),
      session = gsub(".*(session_)", "\\1", comment)
    )

# Fit a logarithmic model
model = lm(log(accel_g) ~ log10(base_weight), data = df)

# Create new data for predictions
new_data = data.frame(
  base_weight = seq(15, 150, 5)
)

# Generate predictions and prediction intervals
au = augment(model, newdata = new_data, interval = "prediction") %>% 
  mutate(across(c(.fitted:.upper), exp))

# Plot the data and model
df %>% 
  ggplot() + 
  geom_point(aes(base_weight, accel_g)) +
  geom_line(data = au, aes(base_weight, y = .fitted)) + 
  geom_ribbon(data = au, aes(base_weight, ymin = .lower, ymax = .upper), alpha = 0.3) +
  geom_hline(yintercept = 1) + 
  labs(x = "Weight Lifted (lbF)", y = "Force Multiplier") + 
  coord_cartesian(x = c(20, 130))
```

# Tracking Progress

The key reason for plotting and modeling this data is to establish a
**baseline** for power across multiple weights. This baseline enables me
to: 1. Identify my current ability to apply force at various weights. 2.
Track changes in power over time.

By training power, I hope to shift this curve upwards—indicating
improvements in my ability to apply more force in shorter periods. The
improvements may occur uniformly across all weights or be more
pronounced at specific intervals. Either way, this approach gives me a
measurable way to assess progress.

# Conclusion

This process highlights the importance of establishing a baseline for
any targeted training program. By visualizing and modeling the
force-velocity relationship, I have a clear starting point to monitor
how training affects my power output.

Ultimately, the goal is not just to collect data but to use it as a
foundation for **meaningful improvement**. Over time, I’ll revisit this
curve to measure the impact of my training efforts.


# Execution (In Practice)

So what was my process?

I used my **Tindeq** to establish my power output (in G's) across a range of different weights.

Again, by lifting it was fast as possible while maintiang good grip form, I could measure  maximum force I could apply during each lift. 

Here’s how I conducted this process:

### Steps for Measuring Power
1. **Setup**:
   - Over several sessions, I selected a random assortment of weights ranging from 25 to 130 lbs.
   - Using a **half-crimp grip**, I connected the Tindeq to the weight and ensured it was secure.

2. **Grip and Position**:
   - I used a half-crimp grip for consistency, focusing on preventing my fingers from uncurling during the lift. If my fingers uncurl, it:
     - Changes the grip position, affecting the measurement.
     - Results in inaccurate force data, as my body’s movement compensates unnaturally.
   - I positioned myself in a **half-squat stance**, with the weight lifted from a low position close to the floor upwards.

3. **Measuring Base Weight**:
   - Before each lift, I slowly raised the weight to ensure I had an accurate reading of its true weight. 
   - This was recorded as the **base weight**.

4. **Peak Force Measurement**:
   - For each lift, I explosively lifted the weight upwards as quickly as possible while maintaining consistent grip and form. 
   - The Tindeq captured the **peak force** during the lift.
   - I performed this process for both my left and right hands individually, recording data for each.

5. **Data Export**:
   - After each session, I exported the data from the Tindeq app, including base weight and peak force measurements.

### Why Modeling Matters
Once I collected this data, I plotted it to visualize the relationship between base weight and acceleration (expressed in G's as \( \text{peak force} / \text{base weight} \)). To analyze this trend, I fit a **logarithmic model** to the data.

The model helps me:
1. **Understand the Baseline**:
   - It shows how my acceleration diminishes as weight increases, which is expected based on the physics of force and mass.
2. **Track Progress**:
   - By re-measuring in the future, I can see if the curve shifts upwards, indicating improved ability to apply force across all weights or specific parts of the curve.

For readers interested in the technical implementation, here’s the R code I used to fit the model and generate predictions:




