---
title: Kara `r emo::ji("heart")` Elon
author: 'Mike Gaunt'
date: '2023-11-14'
slug: kara-heart-elon
categories: ["R", "Text Analysis"]
tags: ["fun", "strings", '5Min Tutorials', 'web scraping']
description: "Lets see how deep that love runs."
image: ~
math: ~
license: ~
hidden: no
comments: yes
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(cache = F, warning = FALSE, message = FALSE
                      ,dpi = 300, cache.lazy = FALSE,
                      tidy = "styler", fig.width = 4, fig.height = 3)
```

```{r echo=FALSE, message=FALSE, eval=T}
#run this if you download the rds object and have it saved locally
data_pro = readRDS(
  here::here("content/post/2023-11-14-kara-heart-elon", "kara_heart_elon_pro.rds")
)


```

# Intro

Kara loves to talk about Elon Musk. 

A couple of friends asked me how often I thought Kara directly mentions/talks about Elon Musk 

This is my attempt to quantify that.

# Method

The data used for this was copied directly from the Apple Podcast App website. 

I only use show note descriptions so any reference to Elon in a podcast but not in the show notes will not be counted.

```{r echo=F, message=FALSE}
require(tidyverse) #entire tidyverse package suite - including purrr, dplyr, ggplot, etc
require(magrittr) #for code piping
require(lubridate) #for working with dates and datetimes
library(plotly)
library(crosstalk)
```

# Overview 

Monthly frequency of tech personalities direct mentions in Pivot podcast descriptions.

```{r echo=FALSE}
person_counts = c("Elon", "Bezos", "Zuckerberg") %>% 
  map_df(~{
    data_pro %>% 
      mutate(flag_detect = str_detect(desc, .x)) %>% 
      mutate(variable = .x)
  })
```

```{r echo=FALSE}
temp_data = person_counts %>% 
  mutate(floor_month = lubridate::floor_date(date, "month")
         ,count = 1) %>% 
  gauntlet::count_percent_zscore(
    grp_c = c("floor_month", "variable", "flag_detect")
    ,grp_p = c("floor_month", "variable")
    ,col = count
    ,rnd = 2
  ) %>% 
  filter(flag_detect) %>% 
  complete(floor_month, variable, flag_detect, fill = list(count = 0, percent = 0)) %>% 
  group_by(variable) %>%
  mutate(gauntlet::auto_make_rMean_col(col = count, width = 3)) %>% 
  ungroup()

temp_data_sh = SharedData$new(temp_data)

bscols(
  widths = c(12)
       ,filter_select(id = "id_1_zz",label = "Pick an Asshole: ", sharedData = temp_data_sh, multiple = T, group = ~variable)
       ,subplot(
         temp_data_sh %>% 
           plot_ly(x = ~floor_month, y = ~count, color = ~variable, legendgroup = ~variable
                   ,type="scatter", mode="lines+markers", showlegend = T) %>% 
           # add_trace(x = ~floor_month, y = ~count_rMean_3, color = ~variable, legendgroup = ~variable
                   # ,type="scatter", mode="lines", showlegend = T) %>%
           layout(yaxis = list(title = "Count"), xaxis = list(title = "Year"))  
         ,temp_data_sh %>% 
           plot_ly(x = ~floor_month, y = ~percent, color = ~variable #change
                   ,type = "bar", legendgroup = ~flag_detect , showlegend = F) %>% 
           
           layout(yaxis = list(title = "Percent"), xaxis = list(title = "Year")), nrows = 2
         ,shareX = T, shareY = T, heights = c(.6,  .4)
       ))
```

Specifically Elon and Elon affiliated stuff.  `r emo::ji("rocket")` `r emo::ji("rocket")` `r emo::ji("rocket")`

```{r echo=FALSE}
person_counts = c("Elon", "Twitter", "Elon|Twitter", "SpaceX", "Tesla") %>% 
  map_df(~{
    data_pro %>% 
      mutate(flag_detect = str_detect(desc, .x)) %>% 
      mutate(variable = .x)
  })
```

```{r echo=FALSE}
temp_data = person_counts %>% 
  mutate(floor_month = lubridate::floor_date(date, "month")
         # ,count = as.numeric(flag_detect)
         ,count = 1) %>% 
  gauntlet::count_percent_zscore(
    grp_c = c("floor_month", "variable", "flag_detect")
    ,grp_p = c("floor_month", "variable")
    ,col = count
  ) %>% 
  filter(flag_detect)

temp_data_sh = SharedData$new(temp_data)


bscols(
  widths = c(12)
       ,filter_select(id = "id_1_zz",label = "Pick a Variable: ", sharedData = temp_data_sh, multiple = T, group = ~variable)
       ,subplot(
         temp_data_sh %>% 
           plot_ly(x = ~floor_month, y = ~count, color = ~variable, legendgroup = ~variable
                   ,type="scatter", mode="lines+markers", showlegend = T) %>% 
           # add_trace(x = ~floor_month, y = ~count_rMean_3, color = ~variable, legendgroup = ~variable
                   # ,type="scatter", mode="lines", showlegend = T) %>%
           layout(yaxis = list(title = "Count"), xaxis = list(title = "Year"))  
         ,temp_data_sh %>% 
           plot_ly(x = ~floor_month, y = ~percent, color = ~variable #change
                   ,type = "bar", legendgroup = ~flag_detect , showlegend = F) %>% 
           
           layout(yaxis = list(title = "Percent"), xaxis = list(title = "Year")), nrows = 2
         ,shareX = T, shareY = T, heights = c(.6,  .4)
       ))
```












