---
title: "5 Minute Tutorials: Using Protocol Buffers in R"
author: 'Mike Gaunt'
date: '2025-01-28'
slug: quickie-protobuf
categories: ["R", "Programming"]
tags: ["protobuf", "GTFS", "5Min Tutorials"]
description: "Quick tutorial on how to use Protocol Buffers in R with GTFS-realtime data."
image: ~
math: ~
license: ~
hidden: no
comments: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
full_path = here::here("content/post/2025-01-29-quickie-proto")

library(tidyverse)
library(httr)
library(RProtoBuf)
library(jsonlite)
```

# Quick Overview

Hi hi!  
This is a quick tutorial on using **Protocol Buffers** (protobufs) in R, particularly for working with **GTFS-realtime** data. If you've ever struggled to parse protobuf data, this guide will walk through the key steps to get you started.

## What are Protocol Buffers?

Protocol Buffers (protobufs) are a language-neutral, platform-neutral extensible mechanism for serializing structured data. They are commonly used in real-time data applications, such as **GTFS-realtime**, because they are lightweight and efficient. Unlike JSON or XML, which store data in human-readable formats, protobufs store data in binary format, making them compact and fast to transmit. 

In R, we use the **RProtoBuf** package to read and process protobufs.

# Getting Started

### Checking Available Protobuf Descriptors

A **descriptor pool** is a collection of message types and enums available in a protobuf schema. In R, we can inspect available descriptors with:

```{r}
ls("RProtoBuf:DescriptorPool")
```

Descriptors define the structure of the data. Think of them like a schema for a database or the column definitions in a data frame.

### Downloading and Reading the GTFS-realtime Protobuf

GTFS-realtime data uses a predefined protobuf schema. First, download the `.proto` definition file, and then load the protobuf schema into R:

```{r echo=F, message=F}
proto_file = "https://gtfs.org/documentation/realtime/gtfs-realtime.proto"
download.file(proto_file, destfile = here::here(full_path, "gtfs-realtime.proto"))
readProtoFiles("gtfs-realtime.proto")
```

```{r eval=F}
proto_file = "https://gtfs.org/documentation/realtime/gtfs-realtime.proto"
download.file(proto_file, destfile = "gtfs-realtime.proto")
readProtoFiles("gtfs-realtime.proto")
```

```{r}
ls("RProtoBuf:DescriptorPool")
```

This lets us see the available message types defined in the GTFS-realtime schema.

### Understanding a Protobuf Descriptor

A **descriptor** is like a class definition for a protobuf message. Here’s how we inspect one:

```{r}
descriptor = RProtoBuf::P("transit_realtime.FeedMessage")
descriptor
```

Descriptors define the message structure, including nested message types. In R, they may look similar to functions or methods, but they represent structured data rather than executable code.

# Loading a GTFS-realtime Feed

GTFS-realtime feeds are updated frequently and are stored in **binary format** for efficiency. We need to download and parse them properly.

First, retrieve the binary protobuf feed:

```{r}
feed_url <- "http://s3.amazonaws.com/commtrans-realtime-prod/tripupdates.pb"
resp = httr::GET(feed_url)
feed_data = resp %>% content("raw")
```

Now, parse the binary data using the descriptor:

```{r}
gtfs_feed = P("transit_realtime.FeedMessage")$read(feed_data)
```

The `$read()` function is similar to a method in Python—it’s how we deserialize binary data into a structured format.

### Converting the Feed to JSON

To work with the data more easily, convert it to JSON:

```{r}
gtfs_feed_json = gtfs_feed$toJSON()
gtfs_feed_json = jsonlite::fromJSON(gtfs_feed_json, simplifyVector = FALSE)
```

Now, let’s extract and inspect the entities in the feed:

```{r}
gtfs_feed_json %>% 
  .[['entity']] %>% 
  enframe() %>% 
  unnest_wider(value) %>% 
  unnest_wider(tripUpdate)
```

# Final Thoughts

This tutorial showed how to:
- Load and inspect GTFS-realtime protobuf schemas
- Download and parse GTFS-realtime data
- Convert protobufs to JSON for easier manipulation

Protobufs are an efficient way to handle structured real-time data, and **RProtoBuf** makes it possible to work with them in R.

Thanks for reading, and happy coding!
